Gammapy: A Python package for gamma-ray astronomy

### 2023-08-04 src/ms-review.tex
Further language edits

\usepackage{graphicx}
\usepackage{url}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{import}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[autostyle]{csquotes}
\usepackage{siunitx}

@@

\begin{document}
\newcommand{\PythonUrl}{\url{http://fits.gsfc.nasa.gov/}\xspace}
\newcommand{\FitsUrl}{\url{http://fits.gsfc.nasa.gov/}\xspace}
\newcommand{\GammapyUrl}{\url{http://gammapy.org}\xspace}
\newcommand{\GadfUrl}{\url{http://gamma-astro-data-formats.readthedocs.io/}\xspace}
\newcommand{\ReadthedocsUrl}{\url{https://readthedocs.org/}\xspace}
\newcommand{\TravisUrl}{\url{https://www.travis-ci.org/}\xspace}

@@

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{\textit{Fermi}-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

@@

\abstract
	{
		Traditionally, TeV-\gammaray astronomy has been conducted
		by experiments employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the Cherenkov Telescope Array Observatory (CTAO), will be operated as open observatories.
		Alongside the data, they will also make the associated software tools available to a wider community.
    This necessity prompted the development of open, high-level, astronomical software customised for high-energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of its first long-term-support release, version 1.0.
		Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialised data products that
		are interoperable with other software packages.
	}
	{
		Starting from event lists and instrument response functions,
		\gammapy provides functionalities to reduce these data by binning them in energy and sky coordinates.
		Several techniques for background estimation are implemented in the package to handle the residual hadronic background affecting \gammaray instruments.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show, using publicly-available gamma-ray data,
		the capabilities of \gammapy in multiple traditional and novel \gammaray analysis scenarios,
		such as spectral and spectro-morphological modelling and estimations of a spectral energy
		distribution and a light curve.
		Its flexibility and its power are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model.
	}{}

@@

% Main part
\section{Introduction}
\label{sec:introduction}
%
\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/big-picture.pdf}
	\caption{
		Core idea and relation of \gammapy to different \gammaray instruments
		and the gamma astro data format (GADF). The top left shows the
		group of current and future pointing instruments based on the 
		imaging atmospheric Cherenkov technique (IACT). This includes
		instruments such as the Cherenkov Telescope Array Observatory (CTAO),
		the High Energy Stereoscopic System (H.E.S.S.), the
		Major Atmospheric Gamma Imaging Cherenkov Telescopes (MAGIC),
		and the Very Energetic Radiation Imaging Telescope Array System (VERITAS).
		The lower left shows the group of all-sky instruments such as the
		Fermi Large Area Telescope (\textit{Fermi}-LAT) and the High Altitude
		Water Cherenkov Observatory (HAWC). The calibrated data of all those
		instruments can be converted and stored into the common GADF data format, that \gammapy can read.
		The \gammapy package is a community-developed project that
		provides a common interface to the data and analysis of all
		these \gammaray instruments, allowing users to easily combine data from
		different instruments and perform joint analyses.
		\gammapy is built on the scientific Python ecosystem, and the required dependencies
		are shown below the \gammapy logo.
	}
	\label{fig:big_picture}

\end{figure*}
%
%% \gammaray astronomy
Modern astronomy offers the possibility to observe and study astrophysical sources across
all wavelengths. The \gammaray range of the electromagnetic spectrum 
provides us with insights into the most energetic processes in the universe such as
those accelerating particles in the surroundings of black holes or remnants of
supernova explosions.

In general, \gammaray astronomy relies on the detection of individual 
photon events and reconstruction of their incident direction as well
as energy. As in other branches of astronomy, this can be
achieved by satellite as well as ground-based \gammaray instruments.
Space-borne instruments such as the Fermi Large Area Telescope (LAT)
rely on the pair-conversion effect to detect \gammarays and track
the positron-electron pairs created in the detector to reconstruct the incident direction 
of the incoming \gammaray. The energy of the photon is estimated using a calorimeter
at the bottom of the instrument. The energy range of instruments like \fermi \citep{Atwood2009}, referred to as
high-energy (HE), goes approximately from tens of \si{MeV} to hundreds of \si{GeV}.

Ground-based instruments, instead, use the Earth's atmosphere as a particle detector, relying on the effect that
cosmic \gammarays interacting in the atmosphere create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on the observation of these extensive air showers to estimate the
primary \gammaray photons incident direction and energy.
These instruments operate in the so-called very-high-energy (VHE) regime,
covering the energy range from few tens of \si{GeV} up to the \si{PeV}.
There are two main categories of ground-based instruments.

\begin{itemize}
\setlength\itemsep{1em}

\item \textit{Imaging Atmospheric Cherenkov Telescopes (IACTs)} obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by charged particles in the cascade and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

\item \textit{Water Cherenkov Detectors (WCDs)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV, and large duty-cycle, but a higher energy threshold and
lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

\end{itemize}

%% Context
While \fermi data and analysis tools have been public from early years
of the project \citep{Atwood2009}, ground-based \gammaray astronomy has been historically conducted
through experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been so far successful, it does not
permit easy combination of data from several instruments. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially in light of the fact that different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of energy range covered,
duty cycle and spatial resolution.

%TODO: better transition into context
The Cherenkov Telescope Array Observatory (CTAO) will be the first ground-based
\gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by \cta will be reconstructed event lists and corresponding instrument response functions.}
will be shared publicly after some proprietary period, along with the software required to analyze them.
The adoption of an open and standardised \gammaray data format, beside being a necesity for future observatories such as \cta,
will be greatly beneficial to the current generation of instruments, eventually allowing to exploit their data legacies even
beyond the end of their scientific operations.

%% Context : formats
The usage of a common data format is facilitated by the remarkable similarity of the data reduction workflow of all \gammaray telescopes.
After data calibration, shower events are reconstructed and
a gamma / hadron separation is applied to reject cosmic-ray-intiated showers and build lists of \gammaray-like events.
The latter can then be used, taking into account the observation-specific instrument response functions (IRFs),
to derive scientific results, such as spectra, sky maps or light curves.
Once the data is reduced to a list of events with reconstructed physical properties of the primary particle,
the information is independent of the data-reduction process, and, eventually, of the detection technique. This implies,
for example, that high-level data from IACTs and WCDs can be represented
with the same data model.
The efforts to create a common format usable by various instruments
converged in the so-called \textit{Data Formats for \gammaray Astronomy}
initiative~\citep{gadf_proc,gadf_universe}, abbreviated to
\texttt{gamma-astro-data-formats} (\gadf). This community-driven intitiative proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of \gammaray-like events with their reconstructed and observed
quantities such as energy, incident direction and arrival time and a parametrisation of
the IRFs associated with the event list data. %% where? in the data units of a FITS file.

%% Context: development of open software
In the past decade \textit{Python} has become extremely popular as a scientific programming language,
in particular in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a \enquote{glue} language between different programming languages and last but not least
the rich ecosystem of packages and its open and supportive community \citep{Momcheva2015}.
In the sub-field of computational astronomy, the \astropy project \citep{astropy} was created in 2012
to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units
as well as reading and writing FITS files.

%% \gammapy: concept and goals
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. The various \gammaray instruments
can export their data to a common format (\gadf) and then combine and analyse
them using a common software library.
The \gammapy package is an independent community-developed software project.
It has been selected to be the core library for the Science Analysis tools of \cta,
but also involves contributors associated to other instruments.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

With the public availability of the \gadf format specifications and the
\gammapy package, some experiments started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf format \citep{HESS_DR1} for data level 3 (DL3) \gammaray data.
This data release served as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.][]{Mohrmann2019}. Two observations of the Crab nebula have
been released by the \magic collaboration~\citep{magic_performance}.
Using these public data from \fermi, \hess, \magic and additional observations
provided by FACT and \veritas, the authors of \cite{joint_crab} presented
a combined analysis of \gammaray data from different instruments for the first time.
Later the \hawc collaboration also released a limited test dataset of the Crab Nebula,
which was used to validate the \gammapy package in \cite{Olivera2022}.

The increased availability of public data that followed the definition of
a common data format, and the development of \gammapy as a community-driven open software,
led the way toward a more open science in the very-high-energy \gammaray astronomy domain.
The adoption of \gammapy as science tools strengthens the commitment of the future \cta Observatory to the FAIR principles
(Findable, Accessible, Interoperable and Reusable) principles \citep{FAIR16, FAIR22} 
that define the key requirements for open science.

%% Paper outline
In this article, we describe the general structure of the \gammapy package,
its main concepts, and organisational structure. We start in
Section~\ref{sec:gammaray-data-analysis} with a general overview
of the data analysis workflow in very high-energy \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package 
in Section~\ref{sec:gammapy-package}, while also
describing the various subpackages it contains. Section~\ref{sec:applications}
presents a number of applications, while Section~\ref{sec:gammapy-project}
finally discusses the project organization.

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and names given as the corresponding Python code suffix, e.g. 
		\code{gammapy.data}. Below each icon there is a list of the most
        important objects defined in the sub-package. The light grey folder
		icons show the subpackages for the most fundamental data structures such 
		as maps and IRFs. The bottom of the figure shows the high-level analysis
		sub-module with its dependency on the YAML file format. 
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
The data analysis process in \gammaray astronomy is usually split into two stages.
The first one deals with the data processing from detector measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

The second stage, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level science products
such as spectra, light curves and catalogs. The core of the analysis consists in predicting
the result of an observation by modelling the flux distribution of an astrophysical
object and pass it through a model of the instrument. The methods applied here are more
generic and are broadly shared across the field. The similarity in the high-level
analysis conceptionally also allow for easily combining data from multiple instruments.
This part of the data analysis is supported by \gammapy.

\subsection{DL3: Events And Instrument Response Functions}
An overview of the typical steps in the high level analysis is shown in the upper
row of Figure~\ref{fig:data_flow}. The high level analysis starts at the DL3 
data level, where \gammaray data is represented as lists of \gammaray-like events and their
corresponding IRFs $R$ and ends at the DL5/6 data level, where the physically relevant 
quantities such as fluxes, spectra and light curves of sources have been derived. DL3 data are typically 
bundled into individual observations, corresponding to
stable periods of data acquisition. For IACT instruments, for which the \gadf
data model and \gammapy were initially conceived, this duration is typically 
$t_{\rm obs}=15 - 30\,{\rm min}$. Each observation is assigned a unique integer ID for reference.
The event list is just a simple table with one event per row and
the measured event properties as columns. These properties e.g., include reconstructed
incident direction and energy, arrival time and reconstruction quality.

@@

\begin{equation}
	\begin{split}
   R(p, E|p_{\rm true}, E_{\rm true}) = &~A_{\rm eff}(p_{\rm true}, E_{\rm true}) \\
	& \cdot PSF(p|p_{\rm true}, E_{\rm true})\\
    & \cdot E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
	\end{split}
\end{equation}

\noindent Where:
\begin{itemize}
\setlength\itemsep{1em}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability density of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments typically consider radial symmetry of the PSF. With this assumption the probability density 
  $PSF(\Delta p|p_{\rm true}, E_{\rm true})$ only depends on the angular separation between true
 and reconstructed direction defined by $\Delta p = p_{\rm true} - p$.  
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$, the probability density of the event
  migration, $\mu=\frac{E}{E_{\rm true}}$.
\end{itemize}

In addition to the instrument characteristics described above, there is also the instrumental
background, that results from hadronic events being misclassified as \gammaray events. These
events constitute a uniform background to the \gammaray events. For \fermi the residual hadronic background
is very small ($<1\%$) because of its veto layer and can often be neglected. For IACTs and WCDs in contrast,
the background can account for a very large part ($>95\%$) of the events and must be treated
accordingly in the analysis. As the background is very specific to the instrument, \gammapy typically 
relies on the background models provided with the DL3 data. The background usually only depends
on the reconstructed event position and energy $Bkg(p, E)$. However in general all IRFs depend on the
geometrical parameters of the detector, e.g. location of an event in the FoV or the elevation
angle of the incoming direction of the event. Consequently IRFs might be parametrised as functions
of detector specific coordinates as well.

@@

\subsection{From DL3 to DL4: Data Reduction}
The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of spatial dimensions, which either define 
a spherical projection from celestial coordinates to a pixelised image space
or a single region on the sky.

>>>>

\usepackage{graphicx}
\usepackage{txfonts,textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{import}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[autostyle]{csquotes}
\usepackage{siunitx}

@@

\newcommand{\PythonUrl}{\url{http://fits.gsfc.nasa.gov/}\xspace}
\newcommand{\FitsUrl}{\url{http://fits.gsfc.nasa.gov/}\xspace}
\newcommand{\GammapyUrl}{\url{http://gammapy.org}\xspace}
\newcommand{\GadfUrl}{\url{http://gamma-astro-data-formats.readthedocs.io/}\xspace}
\newcommand{\ReadthedocsUrl}{\url{https://readthedocs.org/}\xspace}
\newcommand{\TravisUrl}{\url{https://www.travis-ci.org/}\xspace}

@@

\newcommand{\hess}{H.E.S.S.\xspace}
\newcommand{\hawc}{HAWC\xspace}
\newcommand{\veritas}{VERITAS\xspace}
\newcommand{\magic}{MAGIC\xspace}
\newcommand{\astri}{ASTRI\xspace}
\newcommand{\iact}{IACT\xspace}
\newcommand{\iacts}{IACTs\xspace}
\newcommand{\cta}{CTA\xspace}
\newcommand{\swgo}{SWGO\xspace}
\newcommand{\irf}{IRF\xspace}
\newcommand{\irfs}{IRFs\xspace}
\newcommand{\fermi}{Fermi-LAT\xspace}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\gammarays}{$\gamma$-rays\xspace}
\newcommand{\gadf}{GADF\xspace}
\newcommand{\milagro}{MILAGRO\xspace}
\newcommand{\github}{GitHub\xspace}

\begin{document}

@@

\abstract
        {
                Traditionally, TeV-\gammaray astronomy has been conducted
                by experiments employing proprietary data and analysis software.
                However, the next generation of \gammaray instruments,
                such as the Cherenkov Telescope Array Observatory (CTAO), will be operated as open observatories.
                Alongside the data, they will also make the associated software tools available to a wider community.
    This necessity prompted the development of open, high-level, astronomical software customized for high-energy astrophysics.
        }
        {
                In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
                and illustrate the functionalities of its first long-term-support release, version 1.0.
                Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
                modeling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialized data products that
                are interoperable with other software packages.
        }
        {
                Starting from event lists and instrument response functions,
                \gammapy provides functionalities to reduce these data by binning them in energy and sky coordinates.
                Several techniques for background estimation are implemented in the package to handle the residual hadronic background affecting \gammaray instruments.
                After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
                using Poisson maximum likelihood fitting
                and assuming a variety of spectral, temporal, and spatial models.
                Estimation of flux points, likelihood profiles, and light curves is also supported.
        }
        {
                After describing the structure of the package, we show, using publicly available gamma-ray data,
                the capabilities of \gammapy in multiple traditional and novel \gammaray analysis scenarios,
                such as spectral and spectro-morphological modeling and estimations of a spectral energy
                distribution and a light curve.
                Its flexibility and its power are displayed in a final multi-instrument example,
                where datasets from different instruments, at different stages of data reduction,
                are simultaneously fitted with an astrophysical flux model.
        }{}

@@

% Main part
\section{Introduction}
\label{sec:introduction}
%
\begin{figure*}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/big-picture.pdf}
        \caption{
                Core idea and relation of \gammapy to different \gammaray instruments
                and the gamma astro data format (GADF). The top left shows the
                group of current and future pointing instruments based on the 
                imaging atmospheric Cherenkov technique (IACT). This includes
                instruments such as the Cherenkov telescope array observatory (CTAO),
                the high energy stereoscopic system (H.E.S.S.), the
                major atmospheric gamma imaging Cherenkov telescopes (MAGIC),
                and the very energetic radiation imaging telescope array system (VERITAS).
                The lower left shows the group of all-sky instruments such as the
                Fermi large area telescope (Fermi-LAT) and the high altitude
                water Cherenkov observatory (HAWC). The calibrated data of all those
                instruments can be converted and stored into the common GADF data format, which \gammapy can read.
                The \gammapy package is a community-developed project that
                provides a common interface to the data and analysis of all
                these \gammaray instruments, allowing users to easily combine data from
                different instruments and perform joint analyses.
                \gammapy is built on the scientific Python ecosystem, and the required dependencies
                are shown below the \gammapy logo.
        }
        \label{fig:big_picture}

\end{figure*}
%
%% \gammaray astronomy
Modern astronomy offers the possibility to observe and study astrophysical sources across
all wavelengths. The \gammaray range of the electromagnetic spectrum 
provides us with insights into the most energetic processes in the Universe such as
those accelerating particles in the surroundings of black holes or remnants of
supernova explosions.

In general, \gammaray astronomy relies on the detection of individual 
photon events and reconstruction of their incident direction as well
as energy. As in other branches of astronomy, this can be
achieved by satellite as well as ground-based \gammaray instruments.
Space-borne instruments such as the Fermi Large Area Telescope (LAT)
rely on the pair-conversion effect to detect \gammarays and track
the positron-electron pairs created in the detector to reconstruct the incident direction 
of the incoming \gammaray. The energy of the photon is estimated using a calorimeter
at the bottom of the instrument. The energy range of instruments such as \fermi \citep{Atwood2009}, referred to as
\enquote{high energy} (HE), goes approximately from tens of Megaelectronvolts to hundreds of Gigaelectronvolts.

Ground-based instruments, instead, use Earth's atmosphere as a particle detector, relying on the effect that
cosmic \gammarays interacting in the atmosphere create large cascades of secondary particles, so called \enquote{air showers}, that can be observed from the ground.
Ground-based \gammaray astronomy relies on the observation of these extensive air showers to estimate the
primary \gammaray photons' incident direction and energy.
These instruments operate in the so-called \enquote{very high energy} (VHE) regime,
covering the energy range from a few tens of Gigaelectronvolts up to Petaelectronvolts.
There are two main categories of ground-based instruments.

First there are imaging atmospheric Cherenkov telescopes (IACTs), which obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by charged particles in the cascade and
they use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

Second there are water Cherenkov detectors (WCDs), that detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV, and large duty-cycle, but a higher energy threshold and
lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

%% Context
While \fermi data and analysis tools have been public since the early years
of the project \citep{Atwood2009}, ground-based \gammaray astronomy has been historically conducted
through experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been successful so far, it does not
permit data from several instruments to be combined easily. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially in light of the fact that different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of the energy range covered,
duty cycle, and spatial resolution.

%TODO: better transition into context
The Cherenkov Telescope Array Observatory (CTAO) will be the first ground-based
\gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by \cta will be reconstructed event lists and corresponding instrument response functions.}
will be shared publicly after some proprietary period, along with the software required to analyze them.
The adoption of an open and standardized \gammaray data format, besides being a necessity for future observatories such as \cta,
will be extremely beneficial to the current generation of instruments, eventually allowing their data legacies to be exploited even
beyond the end of their scientific operations.

%% Context : formats
The usage of a common data format is facilitated by the remarkable similarity of the data reduction workflow of all \gammaray telescopes.
After data calibration, shower events are reconstructed and
a gamma-hadron separation is applied to reject cosmic-ray-initiated showers and build lists of \gammaray-like events.
The latter can then be used, taking into account the observation-specific instrument response functions (IRFs),
to derive scientific results, such as spectra, sky maps, or light curves.
Once the data is reduced to a list of events with reconstructed physical properties of the primary particle,
the information is independent of the data-reduction process, and, eventually, of the detection technique. This implies,
for example, that high-level data from IACTs and WCDs can be represented
with the same data model.
The efforts to create a common format usable by various instruments
converged in the so-called {data formats for \gammaray astronomy}
initiative~\citep{gadf_proc,gadf_universe}, abbreviated to
\texttt{gamma-astro-data-formats} (\gadf). This community-driven initiative proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of \gammaray-like events with their reconstructed and observed
quantities such as energy, incident direction and arrival time, and a parameterization of
the IRFs associated with the event list data. %% where? in the data units of a FITS file.

%% Context: development of open software
In the past decade, {Python} has become extremely popular as a scientific programming language,
in particular in the field of data sciences. This success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a \enquote{glue} language between different programming languages, and last but not least
the rich ecosystem of packages and its open and supportive community \citep{Momcheva2015}.
In the subfield of computational astronomy, the \astropy project \citep{astropy} was created in 2012
to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units,
as well as reading and writing FITS files.

%% \gammapy: concept and goals
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. The various \gammaray instruments
can export their data to a common format (\gadf) and then combine and analyze
them using a common software library.
The \gammapy package is an independent community-developed software project.
It has been selected to be the core library for the science analysis tools of \cta,
but it also involves contributors associated with other instruments.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for ND data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimization,
and \matplotlib~\citep{matplotlib} for visualization.

With the public availability of the \gadf format specifications and the
\gammapy package, some experiments have started to make limited subsets of
their \gammaray data publicly available for testing and validating
\gammapy. For example, the \hess collaboration released a limited test
dataset (about 50 hours of observations taken between 2004 and 2008)
based on the \gadf format \citep{HESS_DR1} for data level 3 (DL3) \gammaray data.
This data release served as a basis for validation of open analysis tools, including \gammapy 
\cite[see e.g.,][]{Mohrmann2019}. Two observations of the Crab nebula have
been released by the \magic collaboration~\citep{magic_performance}.
Using these public data from \fermi, \hess, \magic, and additional observations
provided by FACT and \veritas, the authors of \cite{joint_crab} presented
a combined analysis of \gammaray data from different instruments for the first time.
Later the \hawc collaboration also released a limited test dataset of the Crab Nebula,
which was used to validate the \gammapy package in \cite{Olivera2022}.

The increased availability of public data that followed the definition of
a common data format, and the development of \gammapy as a community-driven open software,
led the way toward a more open science in the VHE \gammaray astronomy domain.
The adoption of \gammapy as science tools strengthens the commitment of the future \cta Observatory to the  findable, accessible, interoperable, and reusable (FAIR) principles \citep{FAIR16, FAIR22} 
that define the key requirements for open science.

%% Paper outline
In this article, we describe the general structure of the \gammapy package,
its main concepts, and organizational structure. We start in
Section~\ref{sec:gammaray-data-analysis} with a general overview
of the data analysis workflow in VHE \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package 
in Section~\ref{sec:gammapy-package}, while also
describing the various sub-packages it contains. Section~\ref{sec:applications}
presents a number of applications, while Section~\ref{sec:gammapy-project}
finally discusses the project organization.

\section{Gamma-ray data analysis}
\begin{figure*}[t]
        \centering
        \includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
        \caption{
                \gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and names given as the corresponding Python code suffix, e.g., 
                \code{gammapy.data}. Below each icon there is a list of the most
        important objects defined in the sub-package. The light gray folder
                icons show the sub-packages for the most fundamental data structures such 
                as maps and IRFs. The bottom of the figure shows the high-level analysis
                sub-module with its dependency on the YAML file format. 
    }
        \label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
The data analysis process in \gammaray astronomy is usually split into two stages.
The first one deals with the data processing from detector measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

The second stage, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level science products
such as spectra, light curves and catalogs. The core of the analysis consists in predicting
the result of an observation by modeling the flux distribution of an astrophysical
object and pass it through a model of the instrument. The methods applied here are more
generic and are broadly shared across the field. The similarity in the high-level
analysis conceptionally also allow for easily combining data from multiple instruments.
This part of the data analysis is supported by \gammapy.

\subsection{DL3: Events And Instrument Response Functions}
An overview of the typical steps in the high level analysis is shown in the upper
row of Figure~\ref{fig:data_flow}. The high level analysis starts at the DL3 
data level, where \gammaray data is represented as lists of \gammaray-like events and their
corresponding IRFs $R$ and ends at the DL5/6 data level, where the physically relevant 
quantities such as fluxes, spectra and light curves of sources have been derived. DL3 data are typically 
bundled into individual observations, corresponding to
stable periods of data acquisition. For IACT instruments, for which the \gadf
data model and \gammapy were initially conceived, this duration is typically 
$t_{\rm obs}=15 - 30\,{\rm min}$. Each observation is assigned a unique integer ID for reference.
The event list is just a simple table with one event per row and
the measured event properties as columns. These properties for example, include reconstructed
incident direction and energy, arrival time and reconstruction quality.

@@

\begin{equation}
        \begin{split}
   R(p, E|p_{\rm true}, E_{\rm true}) = &~A_{\rm eff}(p_{\rm true}, E_{\rm true}) \\
        & \cdot PSF(p|p_{\rm true}, E_{\rm true})\\
    & \cdot E_{\rm disp}(E|p_{\rm true}, E_{\rm true})
        \end{split}
,\end{equation}

\noindent where
\begin{itemize}
\setlength\itemsep{1em}
\item $A_{\rm eff}(p_{\rm true}, E_{\rm true})$ is the effective collection area of the detector. It is the product
  of the detector collection area times its detection efficiency at true energy $E_{\rm true}$ and position $p_{\rm true}$.
\item $PSF(p|p_{\rm true}, E_{\rm true})$ is the point spread function (PSF). It gives the probability density of
  measuring a direction $p$ when the true direction is $p_{\rm true}$ and the true energy is $E_{\rm true}$.
  \gammaray instruments typically consider radial symmetry of the PSF. With this assumption the probability density 
  $PSF(\Delta p|p_{\rm true}, E_{\rm true})$ only depends on the angular separation between true
 and reconstructed direction defined by $\Delta p = p_{\rm true} - p$.  
\item $E_{\rm disp}(E|p_{\rm true}, E_{\rm true})$ is the energy dispersion. It gives the probability to
  reconstruct the photon at energy $E$ when the true energy is $E_{\rm true}$ and the true position $p_{\rm true}$.
  \gammaray instruments consider $E_{\rm disp}(\mu|p_{\rm true}, E_{\rm true})$, the probability density of the event
  migration, $\mu=\frac{E}{E_{\rm true}}$.
\end{itemize}

In addition to the instrument characteristics described above, there is also the instrumental
background, that results from hadronic events being misclassified as \gammaray events. These
events constitute a uniform background to the \gammaray events. For \fermi the residual hadronic background
is very small ($<1\%$) because of its veto layer and can often be neglected. For IACTs and WCDs in contrast,
the background can account for a very large part ($>95\%$) of the events and must be treated
accordingly in the analysis. As the background is very specific to the instrument, \gammapy typically 
relies on the background models provided with the DL3 data. The background usually only depends
on the reconstructed event position and energy $Bkg(p, E)$. However in general all IRFs depend on the
geometrical parameters of the detector, such as location of an event in the FoV or the elevation
angle of the incoming direction of the event. Consequently IRFs might be parameterized as functions
of detector specific coordinates as well.

@@

\subsection{From DL3 to DL4: Data Reduction}
The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of spatial dimensions, which either define 
a spherical projection from celestial coordinates to a pixelized image space
or a single region on the sky.

After all data have been projected onto the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
these models can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the FoV background normalization techniques or background measurements
performed for example within reflected regions~\citep{Berge07}.
